---
title: "ToothGrowth Dataset, Basic Inferential Data Analysis"
author: "emeko"
date: "January 25, 2016"
output:
  pdf_document: default
  html_document:
    fig_retina: 2
    fig_width: 6.5
    keep_md: yes
---

# Overview

The *ToothGrowth* data set from the R *datasets* package is a dataset of the response in the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).The data set is provided as a data frame of 60 observations across three variables: length (*len*), supplement (*supp*; a factor that is either 'OJ' for orange juice, or 'VC' for vitamin C (ascorbic acid)), and dose in milligrams (*dose*).

```{r global_options, echo=FALSE}
# Global Setting
# Turn off warning and message to supress package warnings and messages from appearing in the report.
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo = TRUE)
```

```{r, echo=FALSE}
set.seed(66375)
library(datasets)
library(ggplot2)
```

## Basic Exploratory Analysis of the Dataset
The *summary()* function can provide basic information. We know from the description of the data that the tooth length is a reponse (dependent) variable.

```{r}
data('ToothGrowth')

summary( ToothGrowth )
table( ToothGrowth$supp, ToothGrowth$dose )
```

There are **only 10 data points** for each combination of dose and supplement. For this reason,
we will use Student's t-statistic when considering the data.

Since there is nothing in the description of the data that suggests that the observations are paired in any way, we  **assume that the obeservations are not paired**.


# Supplement Dependence

*QUESTION*: Does the observed amount of tooth growth depend on the type of supplement provided?

Lets consider the data faceted by the supplement:

```{r}
ggplot(ToothGrowth, aes(x=len, fill=supp)) +
    geom_histogram( binwidth = 1 ) +
    facet_grid(supp ~ .)
```

From the above plot, the distributions look similar save that the ascorbic acid results appear more spread out.
The difference in spread suggests that the variance for VC-treated cases is greater than for
the OJ cases:

```{r}
aggregate( len ~ supp, ToothGrowth, var)
```

There are large differences in the length variance between the treatments at each dose, which suggests,
in the absence of any prior knowledge that would suggest otherwise, that we should **not treat the
variances as equal**.

The question "Does the observed amount of tooth growth depend on the type of supplement provided?"
suggests a statistical test, where the null-hypothesis ($H_0$) is that the amount of tooth growth is
the same for VC and OJ treated animals ($\mu_{VC} = \mu_{OJ}$) with 95% confidence. Note that there
are multiple dose levels, so we'd like to test the hypothesis at each dosage. The alternate hypothesis is $H_a: \mu_{VC} \ne \mu_{OJ}$.

The length means by dosage and supplement:

```{r}
aggregate( len ~ supp + dose, ToothGrowth, mean)
```

For $H_0: \mu_{VC} = \mu_{OJ}$ at each dose:

```{r}
test.results <- by(
    ToothGrowth,
    ToothGrowth$dose,
    function( data ) {
        t.test( len ~ supp, data, paired = FALSE, var.equal = FALSE )
    }
)

test.results
```

This suggests that at lower doses, we reject the null-hypothesis that true means are equal, but at the
largest does, we cannot reject the null-hypothesis.

Conclusion: at doses of 0.5 and 1.0 milligrams, Orange Juice is better and ascorbic acid at stimulating
tooth growth (with P-values of `r test.results[['0.5']]$p.value` and `r test.results[['1']]$p.value`
repsectively). At 2.0 milligrams, the effect is about the same.

# Supplement Dose

*QUESTION*: Is the observed amount of tooth growth dose dependent?

Lets consider the data faceted by the dose:

```{r}
ggplot(ToothGrowth, aes(x=len, fill=dose)) +
    geom_histogram( binwidth = 1 ) +
    facet_grid(dose ~ .)
```

Visually, the distributions of dose versus response seem to have very different means and similar
variance. There isn't a reason to believe that there exists any difference in variance by dose in
the parent population, so we'll treat the observations as having the same variance.

```{r}
aggregate( len ~ dose, ToothGrowth, var)
```

That being the case, we postulate the null-hypothesis ($H_0$) that there's no difference in the means between doses. The alternative hypothesis is $H_a: \mu_{x} \ne \mu_{y}$ $\forall$ $(x, y) \in [ (0.5, 1), (0.5, 2), (1, 2) ]$.

```{r}
t.test( len ~ dose, subset(ToothGrowth, dose %in% c(0.5, 1)), paired = FALSE)
t.test( len ~ dose, subset(ToothGrowth, dose %in% c(0.5, 2)), paired = FALSE)
t.test( len ~ dose, subset(ToothGrowth, dose %in% c(1, 2)), paired = FALSE)
```

In each of the cases, the means of lengths landed far outside the indicated confidence intervals, and attained a significance level (P-value) of <$1e^{-6}$. Therefore, we can reject the null-hypothesis that the means are equal for the alternative hypothesis that they are not equal.

Conclusion: tooth growth in guinea pigs has a dose-dependent response to vitamin C.


## Appendix

1. R libraries used 
        - datasets, ggplot2
        
2. ToothGrowth R documentation - ?ToothGrowth
        
3. TeX is required to create the PDF output - You should install a recommended TeX distribution for your platform

4. Source code - https://github.com/emeko/StatInfProject